Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-07-13 write-up/covid-writeup.tex
Update covid-writeup.tex

A molecular test conducted on an infected individual may return a negative result.  Such false negatives are very strongly associated with when the test is conducted.  In the incubation phase, most molecular tests will return a false negative result.  Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility~\citep{Mina2020}.   Moreover, most molecular tests are performed via nasopharyngeal swab.  Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.   Systematic reviews suggest that 70\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020,Cohen2020}.

@@

\begin{figure}[!th]
\centering
\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.30$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
%\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{figure}

@@

The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.  This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.  Therefore, the IPW estimates accounting for measurement-error are $5.6\%$ and $4.2\%$ respectively.

@@

The unweighted estimate for the true prevalence of active COVID-19 disease between April 25th and April 29th is 11.7\%.  Assuming a false negative rate of $30$\% and false positive rate of $2.4$\%, the unweighted estimate is $10.4$\%, leading to an error of $8.5$\%.  Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
\begin{equation*}
\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.085}}{\sigma_Y} = 3.84 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
\end{equation*}
Using inverse-probability weights, the IPW estimates for the true prevalence of active COVID-19 disease between April 25th and April 29th are $7.7$\% and $6.5\%$ respectively.  Under the same sensitivity and specificity, the weighted estimates adjusted for measurement-error are $5.6$\% and $4.2$\%, leading to errors $3.8$\% and $2.4$\% respectively.  Using~\eqref{eq:statdecomp2}, we construct an estimate of the relative sampling rate as follows
\begin{equation*}
\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.038}}{\sigma_Y} = 6.93 \times 10^{-3} \Rightarrow \tilde \Delta = 8.24 \times 10^{-3}  \Rightarrow \tilde M = 3.9,
\end{equation*}
using the Indiana non-random sample.  For the IPW estimate using Facebook data, we have $\tilde M = 3.3$. Therefore, the impact of self selection on bias of $\hat R_t$ appears strong using unweighted data, remaining moderate even using the weighted estimates.

A sensitivity analysis is performed by considering the range of false negative and false positive rates.  Here, sensitivity is assumed to fall between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.39)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses respectively.

@@

Here we extend the IPW analysis to the time-varying setting.  To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.  Recall the COVID-19 testing and positive case counts are known by age, gender, and racial demographics. Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.  As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, Facebook symptom survey and hospitalization data are both used to estimate symptom status per strata.  We do this in two ways which are each detailed below.

@@

Here, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.  This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.  To allow for potential uncertainty, we assume two \emph{synthetic} simple random samples.  The first is a simple random sample of true negative cases denoted $S_{FP}$; the second is simple random sample of true positive cases denoted $S_{FN}$.  As $\mu_t$ and $\theta_{t}$ are estimated independently at each time $t$, we can focus on the estimating equations per time point~$t$ separately.

>>>>

A molecular test conducted on an infected individual may return a negative result.  Such false negatives are very strongly associated with when the test is conducted.  In the incubation phase, most molecular tests will return a false negative result.  Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility~\citep{Mina2020}.   Moreover, most molecular tests are performed via nasopharyngeal swab.  Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.   Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Arevalo2020, Woloshin2020,Cohen2020}.

@@

\begin{figure}[!th]
\centering
\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
%\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{figure}

@@

The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.  This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$.  Therefore, the IPW estimates accounting for measurement-error are $6.2\%$ and $6.1\%$ respectively.

@@

The unweighted estimate for the true prevalence of active COVID-19 disease between April 25th and April 29th is 11.7\%.  Assuming a false negative rate of $13$\% and false positive rate of $2.4$\%, the unweighted estimate is $11.0$\%, leading to an error of $9.2$\%.  Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
\begin{equation*}
\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.092}}{\sigma_Y} = 3.75 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
\end{equation*}
Using inverse-probability weights, the IPW estimates for the true prevalence of active COVID-19 disease between April 25th and April 29th are $7.7$\% and $6.5\%$ respectively.  Under the same sensitivity and specificity, the weighted estimates adjusted for measurement-error are $6.2$\% and $4.9$\%, leading to errors $4.4$\% and $3.1$\% respectively.  Using~\eqref{eq:statdecomp2}, we construct an estimate of the relative sampling rate as follows
\begin{equation*}
\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.044}}{\sigma_Y} = 8.14 \times 10^{-3} \Rightarrow \tilde \Delta = 8.22 \times 10^{-3}  \Rightarrow \tilde M = 3.9,
\end{equation*}
using the Indiana non-random sample.  For the IPW estimate using Facebook data, we have $\tilde M = 3.3$. Therefore, the impact of self selection on bias of $\hat R_t$ appears strong using unweighted data, remaining moderate even using the weighted estimates.

A sensitivity analysis is performed by considering the range of false negative and false positive rates.  Here, sensitivity is assumed to fall between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.44)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses respectively.

@@

The IPW analysis is next extended to the time-varying setting.  To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.  Recall the COVID-19 testing and positive case counts are known by age, gender, and racial demographics. Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.  As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, Facebook symptom survey and hospitalization data are used to estimate symptom status per strata in two ways which are each detailed below.

@@

Here, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.  This corresponds to a false negative rate of $13\%$ and false positive rate of $2.4\%$.  To allow for potential uncertainty, we assume two \emph{synthetic} simple random samples.  The first is a simple random sample of true negative cases denoted $S_{FP}$; the second is simple random sample of true positive cases denoted $S_{FN}$.  As $\mu_t$ and $\theta_{t}$ are estimated independently at each time $t$, we can focus on the estimating equations per time point~$t$ separately.
